Revealing transient strain in geodetic data with Gaussian process regression

### 2017-11-02 RevisedSubmission/hines_hetland_2017_gji.tex
did more

\begin{figure*}
\includegraphics{figures/prior_demo/prior-demo.pdf}
\caption{ 
Panel A) The covariance between $u_i(p)$ and $u_i(p')$,
$C_{u_i}(p,p')$, as a function of the spatial and temporal distance
between $p$ and $p'$. The covariance function $C_{u_i}$ is composed of
squared exponentials to describe the spatial and temporal covariance
of $u_i$ (eq. \ref{eq:ExampleCov}). Panel B) A single realization of
$u_i$ corresponding to the covariance function from Panel A. The
realization is a function of three variables, $t$, $x_\mathrm{e}$, and
$x_\mathrm{n}$, although we only show its dependence on $t$ and one of
the spatial dimensions.
}    
\label{fig:PriorDemo}
\end{figure*}

@@

We then consider the column vector of $n$ GNSS displacement
observations, $\data^*_i$, where the subscript indicates that we are
still only considering component $i$ of displacements. The
observations are made at $m$ stations, and the times and positions for
each observation are described by the set $\points$. $\data^*_i$ can
be considered a realization of the random vector $\data_i$, which is
formed by evaluating eq. (\ref{eq:Data}) at each point in $\points$.
To write out $\data_i$ explicitly, we let $\G$ be an $n \times 6m$
matrix consisting of the basis functions from eq. (\ref{eq:Data})
(i.e., the linear trends and sinusoids for each station) evaluated at
each point in $\points$. The coefficients corresponding to each basis
function are collected into the column vector $\mitbf{m}_i$. The noise
for all the observations are described by the column vector
$\mitbf{\eta}_i$. We can then write $\data_i$ as
\begin{equation}\label{eq:DataVec}
\data_i = u_i(\points) + \mitbf{\eta}_i + \G\mitbf{m}_i,
\end{equation}
where the notation $u_i(\points)$ represents the column vector
$[u_i(p)]_{p \in \points}$.

@@

\begin{figure*}
\includegraphics{figures/context_map/context-map.pdf}
\caption{
Positions of continuous GNSS stations used to estimate transient
strain rates. The colored regions indicate the distribution of seismic
tremor as determined by \citet{Wech2010}. The red dots show the
positions of GNSS stations mentioned in this paper. The blue dot
indicates the location of the transient strain rates shown in Figure
\ref{fig:StrainTs} and the signal-to-noise ratio shown in Figure
\ref{fig:StrainMag}. The blue dashed circle demarcates the spatial
extent of the tremors shown in Figure \ref{fig:StrainMag}.
}    
\label{fig:Context}
\end{figure*}

@@

The distribution of inferred $\alpha$ and $\beta$ are shown in Figure
\ref{fig:NoiseParams}. The estimates of $\beta$ for the easting and
northing components are clustered around 0.5 mm/yr$^{0.5}$. The
corresponding estimates of $\alpha$ tend to cluster around 0
yr$^{-1}$, indicating that the power spectrum of noise indeed decays
proportionally to $f^{-2}$ over a wide band of frequencies. We also
estimate $\alpha$ and $\beta$ for the vertical component of
displacements, with the hope that vertical deformation gradients could
reveal some geophysical signal. For the vertical component, $\beta$ is
relatively large with a median value of 13.5 mm/yr$^{0.5}$. The
inferred values for $\alpha$ are also higher for the vertical
component with a median value of 8.21 yr$^{-1}$. In Figure
\ref{fig:NoiseSamples}, we use the median values of $\alpha$ and
$\beta$ to generate two realizations of FOGM noise for each component.
The realizations span five years, and the easting and northing
realizations drift by about 1 mm over these five years. In the context
of detecting SSEs, which produce several mm's of surface displacement
on the time-scale of weeks, the estimated FOGM noise for the easting
and northing component is negligible. In contrast, the estimated FOGM
noise for the vertical component is larger than the signal we would
expect from SSEs. We suspect that the higher amplitude for the FOGM
noise in the vertical component is accommodating for deficiencies in
our rather simple seasonal model. Based on this analysis, we
henceforth ignore temporally correlated noise in the easting and
northing component because of its low amplitude. We also do not use
vertical displacements because of the presumably low signal-to-noise
ratio.

\begin{figure*}
\includegraphics{figures/noise/noise-params.pdf}
\caption{
Distribution of estimated FOGM parameters (eq. \ref{eq:FOGM}).
parameters are estimated with the REML method for 38 stations in
the Pacific Northwest that are east of $121^\circ$W. ``IQR'' is the
interquartile range.
}   
\label{fig:NoiseParams}
\end{figure*}

\begin{figure*}
\includegraphics{figures/noise/noise-samples.pdf}
\caption{
Two FOGM noise samples for each component. The FOGM parameters
have been set to the median values from Figure \ref{fig:NoiseParams}.
}   
\label{fig:NoiseSamples}
\end{figure*}

@@

\begin{table}\label{tab:Parameters}
\caption{
Optimal values for the parameters in $X_i$ and $T_i$ determined with
the REML method. The covariance function for $T_i$ is indicated by the
``$T_i$'' column. The SE, Wendland, and IBM covariance functions are
defined in eqs. (\ref{eq:TimeSE}), (\ref{eq:Wendland}), and
(\ref{eq:IBM}), respectively. We use eq. (\ref{eq:SE}) for $X_i$.
The parameters are estimated for each of the seven SSEs considered in
this study, and the tabulated values indicate the median and
interquartile ranges of estimates. The ``diff $\log$(REML)'' column
compares the optimal log REML likelihoods to those for when $T_i$ is
the SE covariance function. Negative values indicate that observations
are more consistent with the SE covariance function.
} 
\begin{tabular} {l l l l l l}
$T_i$ & component & $\ell$  & $\phi$   & $\tau$  & diff. $\log$(REML) \\ \hline
SE & east   & 92 $\pm$ 25 km  & 0.62 $\pm$ 0.11 mm  & 0.026 $\pm$ 0.011 yr  &  - \\
SE & north  & 91 $\pm$ 53 km  & 0.43 $\pm$ 0.05 mm  & 0.030 $\pm$ 0.017 yr  &  - \\
Wendland & east   & 95 $\pm$ 30 km  & 0.66 $\pm$ 0.15 mm  & 0.093 $\pm$ 0.044 yr &  0.78 $\pm$ 0.87 \\
Wendland & north  & 92 $\pm$ 57 km  & 0.46 $\pm$ 0.10 mm  & 0.116 $\pm$ 0.057 yr &  0.08 $\pm$ 0.58 \\
IBM & east   & 110 $\pm$ 130 km & 290 $\pm$ 420 mm/yr$^{1.5}$  & -          & -16.4 $\pm$ 7.8 \\
IBM & north  & 150 $\pm$ 560 km & 110 $\pm$ 250 mm/yr$^{1.5}$ & -           & -10.1 $\pm$ 2.3 \\
\end{tabular}
\end{table}

@@

\begin{figure*}
\includegraphics{figures/strain_map/strain-map.pdf}
\caption{
Transient strain rates during the Winter 2015-2016 SSE. The glyphs
show the normal strain rates as a function of azimuth, where orange
indicates compression and blue indicates extension. The shaded regions
indicate the one standard deviation uncertainties for the normal
strain rates. It is common to depict strain rates by showing just the
principal normal strain rates; however, we deviate from this
convention because there is no straight-forward way of showing the
uncertainties for the principal normal strain rates.
}   
\label{fig:StrainMap}
\end{figure*}

@@

\begin{figure*}
\includegraphics{figures/strain_ts/strain-ts.pdf}
\caption{
Three components of the transient horizontal strain rate tensor
estimated at the position shown in Figure \ref{fig:Context}. The
shaded regions indicate one standard deviation uncertainty.
}   
\label{fig:StrainTs}
\end{figure*}

\begin{figure*}
\includegraphics{figures/strain_ts/mag-ts.pdf}
\caption{
(top) Signal-to-noise ratio (eq. \ref{eq:SNR}) at the position shown
in Figure \ref{fig:Context}. (bottom) Frequency of tremors in the
region shown in Figure \ref{fig:Context}.
}   
\label{fig:StrainMag}
\end{figure*}

>>>>

\begin{figure*}
\includegraphics{figures/prior_demo/prior-demo.pdf}
\caption{ 
(A) An example covariance function for transient displacements,
$C_{u_i}(p,p')$, shown as a function of the spatial and temporal
distance between $p$ and $p'$. (B) A single realization of transient
displacements, $u_i(p)$, corresponding to the covariance function from
Panel A. The realization is a function of three variables, $t$,
$x_\mathrm{e}$, and $x_\mathrm{n}$, although we only show its
dependence on $t$ and one of the spatial dimensions.
}    
\label{fig:PriorDemo}
\end{figure*}

@@

We then consider the column vector of $n$ GNSS displacement
observations, $\data^*_i$, where the subscript indicates that we are
still only considering component $i$ of displacements. The
observations are made at $m$ stations, and the times and positions for
each observation are described by the set $\points$. The data vector
$\data^*_i$ can be considered a realization of the random vector
$\data_i$, which is formed by evaluating eq. (\ref{eq:Data}) at each
point in $\points$. To write out $\data_i$ explicitly, we let $\G$ be
an $n \times 6m$ matrix consisting of the basis functions from eq.
(\ref{eq:Data}) (i.e., the linear trends and sinusoids for each
station) evaluated at each point in $\points$. The coefficients
corresponding to each basis function are collected into the column
vector $\mitbf{m}_i$. The noise for all the observations are described
by the column vector $\mitbf{\eta}_i$. We can then write $\data_i$ as
\begin{equation}\label{eq:DataVec}
\data_i = u_i(\points) + \mitbf{\eta}_i + \G\mitbf{m}_i,
\end{equation}
where the notation $u_i(\points)$ represents the column vector
$[u_i(p)]_{p \in \points}$.

@@

\begin{figure*}
\includegraphics{figures/context_map/context-map.pdf}
\caption{
Positions of the GNSS stations used to estimate transient strain
rates. The colored regions indicate the distribution of seismic tremor
as determined by \citet{Wech2010}. The red dots show the positions of
GNSS stations mentioned in this paper. The blue dot indicates the
location of the transient strain rates shown in Figure
\ref{fig:StrainTs} and the signal-to-noise ratio shown in Figure
\ref{fig:StrainMag}. The blue dashed circle demarcates the spatial
extent of the tremors shown in Figure \ref{fig:StrainMag}.
}    
\label{fig:Context}
\end{figure*}

@@

The distribution of inferred values for $\alpha$ and $\beta$ are shown
in Figure \ref{fig:NoiseParams}. The estimates of $\beta$ for the
easting and northing components are clustered around 0.5
mm/yr$^{0.5}$. The corresponding estimates of $\alpha$ tend to cluster
around 0 yr$^{-1}$, indicating that the power spectrum of noise indeed
decays proportionally to $f^{-2}$ over a wide band of frequencies. We
also estimate $\alpha$ and $\beta$ for the vertical component of
displacements, with the hope that vertical deformation gradients could
reveal some geophysical signal. For the vertical component, $\beta$ is
relatively large with a median value of 13.5 mm/yr$^{0.5}$. The
inferred values for $\alpha$ are also higher for the vertical
component with a median value of 8.21 yr$^{-1}$. In Figure
\ref{fig:NoiseSamples}, we use the median values of $\alpha$ and
$\beta$ to generate two realizations of FOGM noise for each component.
The realizations span five years, and the easting and northing
realizations drift by about 1 mm over these five years. In the context
of detecting SSEs, which produce several mm's of surface displacement
on the time-scale of weeks, the estimated FOGM noise for the easting
and northing component is negligible. In contrast, the estimated FOGM
noise for the vertical component is larger than the signal we would
expect from SSEs. We suspect that the higher amplitude for the FOGM
noise in the vertical component is accommodating for deficiencies in
our rather simple seasonal model. Based on this analysis, we
henceforth ignore temporally correlated noise in the easting and
northing component because of its low amplitude. We also do not use
vertical displacements because of the presumably low signal-to-noise
ratio.

\begin{figure*}
\includegraphics{figures/noise/noise-params.pdf}
\caption{
Distribution of inferred values for the parameters in the FOGM noise
model (eq. \ref{eq:FOGM}). ``IQR'' is the interquartile range of
inferred values.
}   
\label{fig:NoiseParams}
\end{figure*}

\begin{figure*}
\includegraphics{figures/noise/noise-samples.pdf}
\caption{
Two samples of FOGM noise for each displacement component. The
parameters for the FOGM noise model have been set to the median values
from Figure \ref{fig:NoiseParams}.
}   
\label{fig:NoiseSamples}
\end{figure*}

@@

\begin{table}\label{tab:Parameters}
\caption{
Optimal values for the parameters in $X_i$ and $T_i$ determined with
the REML method. The covariance function for $T_i$ is indicated by the
``$T_i$'' column. The SE, Wendland, and IBM covariance functions are
defined in eqs. (\ref{eq:TimeSE}), (\ref{eq:Wendland}), and
(\ref{eq:IBM}), respectively. We use eq. (\ref{eq:SE}) for $X_i$.
The parameters are estimated for each of the seven SSEs considered in
this study, and the tabulated values indicate the median and
interquartile ranges of estimated values. The ``diff $\log$(REML)'' column
compares the optimal log REML likelihoods to those for when $T_i$ is
the SE covariance function. Negative values indicate that observations
are more consistent with the SE covariance function.
} 
\begin{tabular} {l l l l l l}
$T_i$ & component & $\ell$  & $\phi$   & $\tau$  & diff. $\log$(REML) \\ \hline
SE & east   & 92 $\pm$ 25 km  & 0.62 $\pm$ 0.11 mm  & 0.026 $\pm$ 0.011 yr  &  - \\
SE & north  & 91 $\pm$ 53 km  & 0.43 $\pm$ 0.05 mm  & 0.030 $\pm$ 0.017 yr  &  - \\
Wendland & east   & 95 $\pm$ 30 km  & 0.66 $\pm$ 0.15 mm  & 0.093 $\pm$ 0.044 yr &  0.78 $\pm$ 0.87 \\
Wendland & north  & 92 $\pm$ 57 km  & 0.46 $\pm$ 0.10 mm  & 0.116 $\pm$ 0.057 yr &  0.08 $\pm$ 0.58 \\
IBM & east   & 110 $\pm$ 130 km & 290 $\pm$ 420 mm/yr$^{1.5}$  & -          & -16.4 $\pm$ 7.8 \\
IBM & north  & 150 $\pm$ 560 km & 110 $\pm$ 250 mm/yr$^{1.5}$ & -           & -10.1 $\pm$ 2.3 \\
\end{tabular}
\end{table}

@@

\begin{figure*}
\includegraphics{figures/strain_map/strain-map.pdf}
\caption{
Transient strain rates during the Winter 2015-2016 SSE. The glyphs
show the normal strain rates as a function of azimuth, where orange
indicates compression and blue indicates extension. The shaded regions
indicate the one standard deviation uncertainties for the normal
strain rates.
}   
\label{fig:StrainMap}
\end{figure*}

@@

\begin{figure*}
\includegraphics{figures/strain_ts/strain-ts.pdf}
\caption{
Components of the transient strain rates at the position shown in
Figure \ref{fig:Context}. The shaded regions indicate the one standard
deviation uncertainties.
}   
\label{fig:StrainTs}
\end{figure*}

\begin{figure*}
\includegraphics{figures/strain_ts/mag-ts.pdf}
\caption{
(top) Signal-to-noise ratio (eq. \ref{eq:SNR}) at the position shown
in Figure \ref{fig:Context}. (bottom) Frequency of seismic tremors in
the region shown in Figure \ref{fig:Context}.
}   
\label{fig:StrainMag}
\end{figure*}
